---
title: "Placenta Metabolomics — Step 0 (Preprocess) & Step 1 (Stats + Outliers)"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
params:
  data_dir: "/Users/atharva/Downloads/Jessica/Raw data"          # change to your folder path when knitting locally
  output_dir: "outputs"  # where CSVs and plots will be written
  demo_file: "mata2022_caco_mar242022.xlsx"
  placenta_file: "MetaPro_placenta_Jan2022.xlsx"
  placenta_norm_sheet: "Normalized data"
  placenta_info_sheet: "Sample Info"
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 5)

# Helper: install if missing
pkg_needed <- c("readxl","dplyr","tidyr","ggplot2","psych","Hmisc","ggpubr")
to_install <- setdiff(pkg_needed, rownames(installed.packages()))
if (length(to_install) > 0) install.packages(to_install, quiet = TRUE)
lapply(pkg_needed, require, character.only = TRUE)

# Paths
data_dir   <- params$data_dir
output_dir <- params$output_dir
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

demo_path     <- file.path(data_dir, params$demo_file)
placenta_path <- file.path(data_dir, params$placenta_file)

norm_sheet <- params$placenta_norm_sheet
info_sheet <- params$placenta_info_sheet
```

# 0. Load raw data

```{r load-data}
# Demographics
data.demo <- readxl::read_excel(demo_path)

# Placenta metabolite matrix + sample info
data.placenta      <- readxl::read_excel(placenta_path, sheet = norm_sheet)
data.placenta.info <- readxl::read_excel(placenta_path, sheet = info_sheet)

# Preview
list(
  demo_cols = colnames(data.demo)[1:12],
  placenta_cols = colnames(data.placenta)[1:12],
  info_cols = colnames(data.placenta.info)
)
```

## 0.1 Keep IDs present in demo (all subjects)

```{r filter-keep}
keep_ids <- data.demo %>%
  dplyr::select(matchid, IDCode)

nrow(keep_ids)
head(keep_ids)
```

## 0.2 Harmonize sample ID -> IDCode (renaming columns)

```{r rename-idcode}
# Per provided script: placenta IDCode comes from Tube Label substring(2, nchar)
data.placenta.info <- data.placenta.info %>%
  dplyr::mutate(
    IDCode = substr(`Tube Label`, 2, nchar(`Tube Label`))
  ) %>%
  dplyr::select(`SAMPLE ID`, IDCode)

# Rename measurement columns (start at column 12 per original)
placenta_fixed <- data.placenta
if (ncol(placenta_fixed) >= 12) {
  meas_cols <- colnames(placenta_fixed)[12:ncol(placenta_fixed)]
  map <- setNames(data.placenta.info$IDCode, data.placenta.info$`SAMPLE ID`)
  new_names <- ifelse(meas_cols %in% names(map), map[meas_cols], meas_cols)
  colnames(placenta_fixed)[12:ncol(placenta_fixed)] <- new_names
}

# Subset to metadata + kept IDs
meta_cols <- c("COMPOUND Name","SUPER META PATHWAY","SUB META PATHWAY","HMDB ID")
present_ids <- intersect(keep_ids$IDCode, colnames(placenta_fixed))
data.placenta.new <- as.data.frame(placenta_fixed[, c(meta_cols, present_ids), drop = FALSE])

dim(data.placenta.new)
head(colnames(data.placenta.new))
```

# 1. Build analysis matrices (subjects as rows)

```{r build-mats}
compound.meta_placenta   <- data.placenta.new[, 1:4, drop = FALSE]
compound.names_placenta  <- make.unique(data.placenta.new$`COMPOUND Name`)
data.mat_placenta        <- data.placenta.new[, -(1:4), drop = FALSE]
rownames(data.mat_placenta) <- compound.names_placenta

data.trans_placenta <- as.data.frame(t(data.mat_placenta), stringsAsFactors = FALSE)
data.trans_placenta$SubjectID <- rownames(data.trans_placenta)
data.trans_placenta <- data.trans_placenta[, c("SubjectID", setdiff(names(data.trans_placenta), "SubjectID"))]

write.csv(data.trans_placenta, file.path(output_dir, "placenta_transposed.csv"), row.names = FALSE)
```

# 2. Missingness overview

```{r missingness}
to_num <- function(v) {
  v <- as.numeric(as.character(v))
  v[!is.finite(v)] <- NA_real_
  v
}

miss_per_met <- sapply(data.trans_placenta[,-1], function(x) mean(is.na(to_num(x))) * 100)
miss_per_sub <- apply(data.trans_placenta[,-1], 1, function(x) mean(is.na(to_num(x))) * 100)

miss_df <- data.frame(
  Feature = names(miss_per_met),
  PctMissing = as.numeric(miss_per_met),
  row.names = NULL,
  check.names = FALSE
) %>% dplyr::arrange(dplyr::desc(PctMissing))

head(miss_df, 10)

write.csv(miss_df, file.path(output_dir, "placenta_missingness_by_metabolite.csv"), row.names = FALSE)
write.csv(
  data.frame(SubjectID = data.trans_placenta$SubjectID, PctMissing = as.numeric(miss_per_sub)),
  file.path(output_dir, "placenta_missingness_by_subject.csv"),
  row.names = FALSE
)

topN <- 30
top_missing <- miss_df %>% head(topN)

ggplot(top_missing, aes(x = reorder(Feature, PctMissing), y = PctMissing)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top Missing Metabolites (Placenta)", x = NULL, y = "% Missing")
```

# 3. Normality (Shapiro–Wilk) per metabolite

```{r shapiro}
compound_placenta.data <- data.trans_placenta[, -1, drop = FALSE]

shapiro.pvals_pl <- sapply(compound_placenta.data, function(x) {
  x <- to_num(x)
  if (length(unique(na.omit(x))) >= 3) {
    as.numeric(tryCatch(shapiro.test(x)$p.value, error = function(e) NA_real_))
  } else NA_real_
})

hist(shapiro.pvals_pl, breaks = 20, main = "Shapiro–Wilk p-values (Placenta)", xlab = "p-value")

shapiro.results_pl <- data.frame(
  Compound = names(shapiro.pvals_pl),
  P_value  = as.numeric(shapiro.pvals_pl),
  Normal   = ifelse(!is.na(shapiro.pvals_pl) & shapiro.pvals_pl >= 0.05, "Yes", "No"),
  stringsAsFactors = FALSE
)

compound.annotation <- compound.meta_placenta
compound.annotation$Compound_Name <- compound.names_placenta
shapiro.results_pl <- merge(shapiro.results_pl, compound.annotation, by.x = "Compound", by.y = "Compound_Name", all.x = TRUE)

head(shapiro.results_pl)

write.csv(shapiro.results_pl, file.path(output_dir, "placenta_shapiro_results.csv"), row.names = FALSE)
```

# 4. Descriptive statistics (per metabolite)

```{r univariate}
mat <- data.trans_placenta[, -1, drop = FALSE]
mat_num <- as.data.frame(lapply(mat, to_num), check.names = FALSE)

keep_cols <- sapply(mat_num, function(v) length(unique(na.omit(v))) >= 3)
mat_num <- mat_num[, keep_cols, drop = FALSE]

placenta_desc <- psych::describe(mat_num)
head(placenta_desc)
write.csv(placenta_desc, file.path(output_dir, "placenta_metabolite_summary_psych.csv"))
```

# 5. Distributions — Histograms with density & normal curve

```{r histograms}
plot_metabolite_distributions <- function(data_trans, pdf_file,
                                          drop_col = "SubjectID",
                                          breaks = 20,
                                          width = 8, height = 6) {
  mat <- data_trans[, setdiff(names(data_trans), drop_col), drop = FALSE]
  grDevices::pdf(pdf_file, width = width, height = height)
  on.exit(grDevices::dev.off(), add = TRUE)

  for (met in names(mat)) {
    x <- as.numeric(as.character(mat[[met]]))
    x[!is.finite(x)] <- NA_real_
    x <- x[is.finite(x)]

    pval <- if (length(unique(na.omit(x))) >= 3) {
      as.numeric(tryCatch(shapiro.test(x)$p.value, error = function(e) NA_real_))
    } else NA_real_

    ptxt <- if (is.na(pval)) "NA (n < 3 unique)" else formatC(pval, format = "g", digits = 3)
    main_title <- paste0("Histogram of ", met, "\nShapiro-Wilk p = ", ptxt)

    hist(x, breaks = breaks, freq = FALSE,
         main = main_title, xlab = "Value",
         col = "lightgray", border = "white")

    if (length(x) > 1 && stats::sd(x, na.rm = TRUE) > 0) {
      lines(stats::density(x, na.rm = TRUE), lwd = 2)
      xfit <- seq(min(x, na.rm = TRUE), max(x, na.rm = TRUE), length.out = 200)
      yfit <- dnorm(xfit, mean = mean(x, na.rm = TRUE), sd = sd(x, na.rm = TRUE))
      lines(xfit, yfit, lwd = 2)
      legend("topright", legend = c("Empirical density","Normal (mu, sigma)"), lwd = 2, bty = "n")
    } else {
      legend("topright", legend = c("Insufficient variation"), bty = "n")
    }
  }
}

plot_metabolite_distributions(
  data.trans_placenta,
  file.path(output_dir, "placenta_histograms_with_normal_shapiro.pdf")
)
```

# 6. Boxplots & Outliers (IQR rule)

```{r boxplots-outliers}
df_long <- data.trans_placenta %>%
  tidyr::pivot_longer(-SubjectID, names_to = "Metabolite", values_to = "Value") %>%
  dplyr::mutate(Value = as.numeric(as.character(Value))) %>%
  dplyr::mutate(Value = ifelse(is.finite(Value), Value, NA_real_))

pdf(file.path(output_dir, "Placenta_Boxplots_One_Per_Page.pdf"), width = 7, height = 5)
unique_mets <- unique(df_long$Metabolite)
for (met in unique_mets) {
  p <- ggplot(dplyr::filter(df_long, Metabolite == met), aes(x = "", y = Value)) +
    geom_boxplot(outlier.color = "red", fill = "lightgreen") +
    labs(title = paste("Boxplot of", met), y = "Value", x = NULL) +
    theme_bw() +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
  print(p)
}
dev.off()

iqr_outliers <- df_long %>%
  dplyr::group_by(Metabolite) %>%
  dplyr::summarise(
    Q1 = quantile(Value, 0.25, na.rm = TRUE),
    Q3 = quantile(Value, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    Lower = Q1 - 1.5*IQR,
    Upper = Q3 + 1.5*IQR
  )

outliers_long <- df_long %>%
  dplyr::left_join(iqr_outliers, by = "Metabolite") %>%
  dplyr::filter(is.finite(Value) & (Value < Lower | Value > Upper)) %>%
  dplyr::select(SubjectID, Metabolite, Value, Lower, Upper, Q1, Q3, IQR)

head(outliers_long)
write.csv(outliers_long, file.path(output_dir, "placenta_outliers_iqr.csv"), row.names = FALSE)
```

# 7. Session info

```{r session}
sessionInfo()
```

**How to use:**  
- In RStudio, open this file, set `params$data_dir` to your folder (e.g., `"~/Downloads/Jessica/Raw data"`), and click **Knit** to HTML.  
- All CSVs and PDFs will be saved under `outputs/`.
